---
title: "Chapter 1: Exploratory Data Analysis"
format: html
---
```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE, fig.align = "center")
```

```{r paged.print=FALSE}
#| cache: FALSE
#| message: FALSE
#| warning: FALSE
library(corrplot)
library(ellmer)
library(gmodels)
library(matrixStats)
library(tidyverse)

set.seed(123)

# Location of the data files. Adjust this path if you keep the data
# files in a different directory.
DATA_DIR <- '../../data'

```
## Exploratory Data Analysis
### Data Dictionaries and Catalogs
```{r paged.print=FALSE}
system_prompt <- "
Improve the attached data dictionary and return it in YAML format. In addition to
the description of features, provide name, description, source, and
size of the dataset.

For each feature, provide the following information:
- feature name
- readable variable name (if required)
- definition of the variable
- data type
- measurement units
- allowed values (for categorical or nominal features, but only if you are sure)
"

data_dictionary <- "
name: ebay Auctions
description: >
    The dataset eBayAuctions.csv contains information on 1972 auctions that
    transacted on eBay.com during May-June 2004.
source: Copyright 2016 Galit Shmueli and Peter Bruce
size:
    observations: 1972
    features: 8
features:
    Category: Category of the auctioned item
    currency: \"US: US dollar, GBP: English pound, EUR: Euro\"
    sellerRating: >
        a rating by eBay, as a function of the number of \"good\" and  \"bad\"
        transactions the seller had on eBay
    Duration: Number of days the auction lasted (set by seller at auction start)
    endDay: Day of week that the auction closed
    ClosePrice: Price item sold at (converted into USD)
    OpenPrice: Initial price set by the seller (converted into USD)
    Competitive?: whether the auction had a single bid (0) or more (1)
"

message <- paste(c(
  "Here is the data dictionary in YAML format:",
  data_dictionary,
  "First ten lines of the dataset in CSV format:",
  read_csv(file.path(DATA_DIR, "eBayAuctions.csv.gz"), show_col_types = FALSE) %>%
    sample_n(10) %>%
    format_delim(delim = ",", na = "")
), collapse = "\n")

chat <- chat_openai(system_prompt, model = "gpt-4o-mini")
response <- chat$chat(message)
```
### Estimates of Location
#### Example: Location Estimates of Population and Murder Rates
```{r paged.print=FALSE}
state <- read_csv(file.path(DATA_DIR, "state.csv"), show_col_types = FALSE)
population <- state[["Population"]]
mean(population)
mean(population, trim = 0.1)
median(population)
```
```{r paged.print=FALSE}
weighted.mean(state[["Murder.Rate"]], w = state[["Population"]])
weightedMedian(state[["Murder.Rate"]], w = state[["Population"]])
```
### Estimates of Variability
#### Example: Variability Estimates of State Population
```{r paged.print=FALSE}
sd(state[["Population"]])
IQR(state[["Population"]])
mad(state[["Population"]])
```
### Exploring the Data Distribution
#### Percentiles and Boxplots
```{r paged.print=FALSE}
quantile(state[["Murder.Rate"]], p = c(0.05, 0.25, 0.5, 0.75, 0.95))
```
```{r paged.print=FALSE}
ggplot(state, aes(y = Population / 1000000)) +
  geom_boxplot(staplewidth = 0.25) +
  labs(y = "Population (millions)") +
  scale_x_discrete(labels = NULL, breaks = NULL)
```
#### Frequency Tables and Histograms
```{r paged.print=FALSE}
breaks <- seq(
  from = min(state[["Population"]]),
  to = max(state[["Population"]]), length = 11)
state %>%
  mutate(BinRange = cut(Population, breaks = breaks,
      right = TRUE, include.lowest = TRUE)) %>%
  count(BinRange, .drop = FALSE)
```
```{r paged.print=FALSE}
ggplot(state, aes(x = Population)) +
  geom_histogram(breaks = breaks, fill = "grey", color = "black") +
  labs(y = "Frequency")
```
#### Density Plots and Estimates
```{r paged.print=FALSE}
ggplot(state, aes(x = Murder.Rate)) +
  geom_histogram(aes(y = after_stat(density)), bins = 12, fill = "grey",
    color = "black") +
  geom_density() +
  labs(x = "Murder Rate (per 100,000)", y = "Density")
```
### Exploring Binary and Categorical Data
```{r paged.print=FALSE}
dfw <- read_csv(file.path(DATA_DIR, "dfw_airline.csv"), show_col_types = FALSE) %>%
  pivot_longer(everything())
ggplot(dfw, aes(x = name, y = value)) +
  geom_col(color = "black", fill = "grey") +
  labs(x = "Cause of delay", y = "Count")
```
### Correlation
```{r paged.print=FALSE}
sp500_px <- read_csv(file.path(DATA_DIR, "sp500_data.csv.gz"), show_col_types = FALSE)
sp500_sym <- read_csv(file.path(DATA_DIR, "sp500_sectors.csv"), show_col_types = FALSE)
etf_symbols <- sp500_sym %>%
  filter(sector == "etf") %>%
  pull(symbol)
etfs <- sp500_px %>%
  filter(Date > "2012-07-01") %>%
  select(all_of(etf_symbols))
corrplot(cor(etfs), method = "ellipse")
```
```{r paged.print=FALSE}
corrplot(cor(etfs), method = "ellipse")
```
#### Scatterplots
```{r paged.print=FALSE}
ggplot(sp500_px %>% filter(Date > "2012-07-01"), aes(x = T, y = VZ)) +
  geom_point(shape = 1) +
  labs(x = "ATT (T)", y = "Verizon (VZ)")
```
### Exploring Two or More Variables
#### Hexagonal Binning and Contours pass:[<span class#"keep-together">(Plotting Numeric-Versus-Numeric Data)</span>]
```{r paged.print=FALSE}
kc_tax <- read_csv(file.path(DATA_DIR, "kc_tax.csv.gz"), show_col_types = FALSE)
kc_tax0 <- kc_tax %>%
  filter(TaxAssessedValue < 750000,
    SqFtTotLiving > 100,
    SqFtTotLiving < 3500)
nrow(kc_tax0)
```
```{r paged.print=FALSE}
ggplot(kc_tax0, (aes(x = SqFtTotLiving, y = TaxAssessedValue))) +
  geom_hex(color = "white") +
  scale_fill_gradient(low = "white", high = "black") +
  labs(x = "Finished Square Feet", y = "Tax-Assessed Value")
```
```{r paged.print=FALSE}
ggplot(kc_tax0, aes(SqFtTotLiving, TaxAssessedValue)) +
  geom_point(shape = 1, alpha = 0.02) +
  geom_density2d(color = "white") +
  labs(x = "Finished Square Feet", y = "Tax-Assessed Value")
```
#### Two Categorical Variables
```{r paged.print=FALSE}
lc_loans <- read_csv(file.path(DATA_DIR, "lc_loans.csv"), show_col_types = FALSE)
x_tab <- CrossTable(lc_loans$grade, lc_loans$status,
  prop.c = FALSE, prop.chisq = FALSE, prop.t = FALSE)
```
```{r paged.print=FALSE}
# Alternative to create crosstable using tidyverse
crosstab <- lc_loans  %>%
  group_by(status, grade) %>%
  count() %>%  # adds column n with the size of each group
  pivot_wider(id_cols = grade, names_from = status, values_from = n)
crosstab
```
#### Categorical and Numeric Data
```{r paged.print=FALSE}
airline_stats <- read_csv(file.path(DATA_DIR, "airline_stats.csv"), show_col_types = FALSE) %>%
  mutate(
    airline = factor(airline,
      levels = c("Alaska", "American", "JetBlue", "Delta", "United",
        "Southwest"))
  ) %>%
  drop_na()

ggplot(airline_stats, aes(x = airline, y = pct_carrier_delay)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 50)) +
  labs(x = "Airline", y = "Daily % of Delayed Flights")
```
```{r paged.print=FALSE}
ggplot(data = airline_stats, aes(airline, pct_carrier_delay)) +
  geom_violin() +
  coord_cartesian(ylim = c(0, 50)) +
  labs(x = "", y = "Daily % of Delayed Flights")
```
#### Visualizing Multiple Variables
```{r paged.print=FALSE}
kc_subset <- kc_tax0 %>%
  filter(ZipCode %in% c(98188, 98105, 98108, 98126))
ggplot(kc_subset, aes(x = SqFtTotLiving, y = TaxAssessedValue)) +
  geom_hex(color = "white") +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Finished Square Feet", y = "Tax-Assessed Value") +
  facet_wrap("ZipCode")
```
# Supplementary Material
## Table 1-2. A few rows of the data.frame state of population and murder rate by state
```{r paged.print=FALSE}
state %>%
  mutate(
    Population = formatC(Population, format = "d", digits = 0, big.mark = ","),
  ) %>%
  head(8)
```
## Table 1-4. Percentiles of murder rate by state
## Table 1-5. A frequency table of population by state
## Table 1-6. Percentage of delays by cause at Dallas/Fort Worth Airport
## Table 1-7. Correlation between telecommunication stock returns
## Figure 1-6. Correlation between ETF returns (Python version)
## Figure 1-6. Correlation between ETF returns using a heat map (R version)
```{r paged.print=FALSE}
sp500_px <- read_csv(file.path(DATA_DIR, "sp500_data.csv.gz"), show_col_types = FALSE)
sp500_sym <- read_csv(file.path(DATA_DIR, "sp500_sectors.csv"), show_col_types = FALSE)
etf_symbols <- sp500_sym %>%
  filter(sector == "etf") %>%
  pull(symbol)
etfs <- sp500_px %>%
  filter(Date > "2012-07-01") %>%
  select(all_of(etf_symbols))
correlation_matrix <- cor(etfs) %>%
  as_tibble() %>%
  mutate(Var1 = colnames(.)) %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "value")

ggplot(correlation_matrix, aes(x = Var1, y = Var2)) +
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2(limits = c(-1, 1))
```
