---
title: "Chapter 3: Statistical Experiments and Significance Testing"
format: html
---
```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE, fig.align = "center")
```

```{r paged.print=FALSE}
#| cache: FALSE
#| message: FALSE
#| warning: FALSE
library(lmPerm)
library(pwr)
library(tidyverse)

set.seed(123)

# Location of the data files. Adjust this path if you keep the data
# files in a different directory.
DATA_DIR <- '../../data'

```
### Resampling
#### Example: Web Stickiness
```{r paged.print=FALSE}
session_times <- read_csv(file.path(DATA_DIR, "web_page_data.csv"), show_col_types = FALSE) %>%
  mutate(Time = Time * 100)

ggplot(session_times, aes(x = Page, y = Time)) +
  geom_boxplot() +
  labs(y = "Time (in seconds)")
```
```{r paged.print=FALSE}
mean_a <- session_times %>% filter(Page == "Page A") %>% pull("Time") %>% mean()
mean_b <- session_times %>% filter(Page == "Page B") %>% pull("Time") %>% mean()
mean_b - mean_a
```
```{r paged.print=FALSE}
perm_fun <- function(x, n_a) {
  n <- length(x)
  x_resampled <- sample(x)
  x_a <- x_resampled[1:n_a]
  x_b <- x_resampled[(n_a + 1):n]
  return(mean(x_b) - mean(x_a))
}
```
```{r paged.print=FALSE}
set.seed(1)
perm_diffs <- rep(0, 1000)
for (i in 1:1000) {
  perm_diffs[i] <- perm_fun(session_times %>% pull(Time), 21)
}
ggplot(tibble(perm_diffs = perm_diffs), aes(x = perm_diffs)) +
  geom_histogram(bins = 12, color = "black", fill = "white") +
  geom_vline(xintercept = mean_b - mean_a, linetype = "dashed") +
  labs(x = "Session time differences (in seconds)", y = "Frequency")
```
```{r paged.print=FALSE}
mean(perm_diffs > (mean_b - mean_a))
```
### Statistical Significance and p-Values
```{r paged.print=FALSE}
obs_pct_diff <- 100 * (200 / 23739 - 182 / 22588)
conversion <- c(rep(0, 45945), rep(1, 382))
perm_diffs <- rep(0, 1000)
for (i in 1:1000) {
  perm_diffs[i] <- 100 * perm_fun(conversion, 23739)
}
ggplot(tibble(perm_diffs = perm_diffs), aes(x = perm_diffs)) +
  geom_histogram(bins = 12, color = "black", fill = "white") +
  geom_vline(xintercept = obs_pct_diff, linetype = "dashed") +
  labs(x = "Conversion rate (percent)", y = "Frequency") +
  annotate("text", x = 0.1, y = 215, label = "Observed\ndifference", size = 3)
```
#### p-Value
```{r paged.print=FALSE}
mean(perm_diffs > obs_pct_diff)
```
```{r paged.print=FALSE}
prop.test(x = c(200, 182), n = c(23739, 22588), alternative = "greater")
```
### t-Tests
```{r paged.print=FALSE}
t.test(Time ~ Page, data = session_times, alternative = "less")
```
### ANOVA
```{r paged.print=FALSE}
four_sessions  <- read_csv(file.path(DATA_DIR, "four_sessions.csv"), show_col_types = FALSE)
summary(aovp(Time ~ Page, data = four_sessions))
```
#### F-Statistic
```{r paged.print=FALSE}
summary(aov(Time ~ Page, data = four_sessions))
```
### Chi-Square Test
#### Chi-Square Test: A Resampling Approach
```{r paged.print=FALSE}
click_rate <-  read_csv(file.path(DATA_DIR, "click_rates.csv"), show_col_types = FALSE)
clicks <- click_rate %>%
  pivot_wider(id_cols = Headline, names_from = Click, values_from = Rate) %>%
  select(Click, `No-click`)
chisq.test(clicks, simulate.p.value = TRUE)
```
#### Chi-Square Test: Statistical Theory
```{r paged.print=FALSE}
chisq.test(clicks, simulate.p.value = FALSE)
```
#### Fisher's Exact Test
```{r paged.print=FALSE}
fisher.test(clicks)
```
### Power and Sample Size
#### Sample Size
```{r paged.print=FALSE}
effect_size <- ES.h(p1 = 0.0121, p2 = 0.011)
pwr.2p.test(h = effect_size, sig.level = 0.05, power = 0.8,
  alternative = "greater")
```
```{r paged.print=FALSE}
effect_size <- ES.h(p1 = 0.0165, p2 = 0.011)
pwr.2p.test(h = effect_size, sig.level = 0.05, power = 0.8,
  alternative = "greater")
```
# Supplementary Material
## Alternative to t-test using statsmodels
## Figure 3-6. Boxplots of the four groups show considerable differences among them
```{r paged.print=FALSE}
graph <- ggplot(four_sessions, aes(x = Page, y = Time)) +
  geom_boxplot() +
  labs(y = "Time (in seconds)") +
  theme_bw()
graph
```
## Visualizing the resampling results for ANOVA
## chi-2 test using sampling with replacement
## Figure 3-7. Chi-square distribution with various degrees of freedom
```{r paged.print=FALSE}
chi <- tibble(x = 1 + (0:99) * (30 - 1) / 99) %>%
  mutate(
    chi_1 = dchisq(x, df = 1),
    chi_2 = dchisq(x, df = 2),
    chi_5 = dchisq(x, df = 5),
    chi_10 = dchisq(x, df = 10),
    chi_20 = dchisq(x, df = 20),
  )
ggplot(chi, aes(x = x)) +
  geom_line(aes(y = chi_1), linetype = 1) +
  geom_line(aes(y = chi_2), linetype = 2) +
  geom_line(aes(y = chi_5), linetype = 4) +
  geom_line(aes(y = chi_20), linetype = 3) +
  labs(x = "Value of chi-square statistic", y = "Probability") +
  annotate("text", 1.5, 0.05, label = "df=1") +
  annotate("text", 3, 0.25, label = "df=2") +
  annotate("text", 6, 0.15, label = "df=5") +
  # annotate("text", 10, 0.1, label = "df=10") +
  annotate("text", 26, 0.05, label = "df=20")
```
## Figure 3-8. Frequency histogram for Imanishi-Kari lab data
```{r paged.print=FALSE}
imanishi <- read_csv(file.path(DATA_DIR, "imanishi_data.csv"), show_col_types = FALSE)
ggplot(imanishi, aes(x = Digit, y = Frequency)) +
  geom_col() +
  scale_x_continuous(breaks = 0:9, labels = 0:9)
```
