---
title: "Chapter 4: Regression and Prediction"
format: html
---
```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE, fig.align = "center")
```
```{r paged.print=FALSE}
#| cache: FALSE
#| message: FALSE
#| warning: FALSE
library(MASS)
library(latex2exp)
library(lubridate)
library(mgcv)
library(splines)
library(tidyverse)

set.seed(123)

# Location of the data files. Adjust this path if you keep the data
# files in a different directory.
DATA_DIR <- '../../data'

```
## Regression and Prediction
### Simple Linear Regression
#### The Regression Equation
```{r paged.print=FALSE}
lung <- read_csv(file.path(DATA_DIR, "LungDisease.csv"))
model <- lm(PEFR ~ Exposure, data = lung)
```
```{r paged.print=FALSE}
PEFR ~ Exposure - 1
```
```{r paged.print=FALSE}
model
```
#### Fitted Values and Residuals
```{r paged.print=FALSE}
fitted <- predict(model)
resid <- residuals(model)
```
### Multiple Linear Regression
#### Example: King County Housing Data
```{r paged.print=FALSE}
house <- read_csv(file.path(DATA_DIR, "house_sales.csv"), show_col_types = FALSE)
head(house[, c("AdjSalePrice", "SqFtTotLiving", "SqFtLot", "Bathrooms",
      "Bedrooms", "BldgGrade")])
```
```{r paged.print=FALSE}
house_lm <- lm(AdjSalePrice ~ SqFtTotLiving + SqFtLot + Bathrooms +
    Bedrooms + BldgGrade,
  data = house, na.action = na.omit)
```
```{r paged.print=FALSE}
house_lm
```
#### Assessing the Model
```{r paged.print=FALSE}
summary(house_lm)
```
#### Model Selection and Stepwise Regression
```{r paged.print=FALSE}
house_full <- lm(AdjSalePrice ~ SqFtTotLiving + SqFtLot + Bathrooms +
    Bedrooms + BldgGrade + PropertyType + NbrLivingUnits +
    SqFtFinBasement + YrBuilt + YrRenovated + NewConstruction,
  data = house, na.action = na.omit)
```
```{r paged.print=FALSE}
step_lm <- stepAIC(house_full, direction = "both")
step_lm
```
#### Weighted Regression
```{r paged.print=FALSE}
house$Year <- year(parse_date_time(house$DocumentDate, "m/d/y"))
house$Weight <- house$Year - 2005
```
```{r paged.print=FALSE}
house_wt <- lm(AdjSalePrice ~ SqFtTotLiving + SqFtLot + Bathrooms +
    Bedrooms + BldgGrade,
  data = house, weight = Weight)
round(cbind(house_lm = house_lm$coefficients,
    house_wt = house_wt$coefficients), digits = 3)
```
### Factor Variables in Regression
#### Dummy Variables Representation
```{r paged.print=FALSE}
head(house[, "PropertyType"])
```
```{r paged.print=FALSE}
prop_type_dummies <- model.matrix(~ PropertyType - 1, data = house)
head(prop_type_dummies)
```
```{r paged.print=FALSE}
lm(AdjSalePrice ~ SqFtTotLiving + SqFtLot + Bathrooms +
    Bedrooms + BldgGrade + PropertyType, data = house)
```
#### Factor Variables with Many Levels
```{r paged.print=FALSE}
table(house$ZipCode)
```
```{r paged.print=FALSE}
zip_groups <- house %>%
  mutate(resid = residuals(house_lm)) %>%
  group_by(ZipCode) %>%
  summarize(med_resid = median(resid), cnt = n()) %>%
  arrange(med_resid) %>%
  mutate(
    cum_cnt = cumsum(cnt),
    ZipGroup = ntile(cum_cnt, 5),
  ) %>%
  select(ZipCode, ZipGroup)
house <- house %>%
  left_join(zip_groups, by = "ZipCode") %>%
  mutate(ZipGroup = as.factor(ZipGroup))
```
```{r paged.print=FALSE}
table(zip_groups[c("ZipGroup")])
```
### Interpreting the Regression Equation
#### Correlated Predictors
```{r paged.print=FALSE}
step_lm$coefficients
```
```{r paged.print=FALSE}
update(step_lm, . ~ . - SqFtTotLiving - SqFtFinBasement - Bathrooms)
```
#### Confounding Variables
```{r paged.print=FALSE}
lm(AdjSalePrice ~ SqFtTotLiving + SqFtLot + Bathrooms +
    Bedrooms + BldgGrade + PropertyType + ZipGroup, data = house,
  na.action = na.omit)
```
#### Interactions and Main Effects
```{r paged.print=FALSE}
lm(formula = AdjSalePrice ~ SqFtTotLiving * ZipGroup + SqFtLot +
    Bathrooms + Bedrooms + BldgGrade + PropertyType, data = house,
  na.action = na.omit)
```
### Regression Diagnostics
#### Outliers
```{r paged.print=FALSE}
house_98105 <- house %>% filter(ZipCode == 98105)
lm_98105 <- lm(AdjSalePrice ~ SqFtTotLiving + SqFtLot + Bathrooms +
    Bedrooms + BldgGrade, data = house_98105)
```
```{r paged.print=FALSE}
sresid <- rstandard(lm_98105)
idx <- order(sresid)
sresid[idx[1]]
```
```{r paged.print=FALSE}
house_98105[idx[1], c("AdjSalePrice", "SqFtTotLiving", "SqFtLot",
    "Bathrooms", "Bedrooms", "BldgGrade")]
```
#### Influential Values
```{r paged.print=FALSE}
df_influential <- tibble(
  std_resid = rstandard(lm_98105),
  cooks_d = cooks.distance(lm_98105),
  hat_values = hatvalues(lm_98105),
)

p <- 6 # Number of model parameters (e.g., intercept + 2 predictors)
D_i <- 0.08 # Target Cook's Distance
cooks_curve_data <- tibble(
  t = seq(-10, 10, length.out = 1000),
  cooks_d = 0.08,
) %>%
  mutate(
    hat_values = 1 / (1 + t^2),
    std_resid = sqrt(D_i * p) * t
  )

ggplot(df_influential, aes(x = hat_values, y = std_resid,
    color = cooks_d > 0.08, shape = cooks_d > 0.08)) +
  geom_path(aes(size = NULL, color = NULL, shape = NULL),
    data = cooks_curve_data, color = "grey", linewidth = 0.5) +
  geom_hline(yintercept = c(-2.5, 2.5), linetype = 2) +
  geom_point(aes(size = 10 * sqrt(cooks_d))) +
  scale_shape_manual(values = c(1, 19)) +
  guides(color = "none", shape = "none", size = "none") +
  coord_cartesian(xlim = c(0, 0.3), ylim = c(-6, 6), expand = FALSE)
```
#### Heteroskedasticity, Nonnormality, and Correlated Errors
```{r paged.print=FALSE}
df <- data.frame(resid = residuals(lm_98105), pred = predict(lm_98105))
ggplot(df, aes(x = pred, y = abs(resid))) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "loess") +
  scale_x_continuous(labels = function(x) format(x, scientific = FALSE))
```
#### Partial Residual Plots and Nonlinearity
```{r paged.print=FALSE}
terms <- predict(lm_98105, type = "terms")
partial_resid <- resid(lm_98105) + terms
```
```{r paged.print=FALSE}
df <- data.frame(SqFtTotLiving = house_98105[, "SqFtTotLiving"],
  Terms = terms[, "SqFtTotLiving"],
  PartialResid = partial_resid[, "SqFtTotLiving"])
ggplot(df, aes(SqFtTotLiving, PartialResid)) +
  geom_point(shape = 1) +
  scale_shape(solid = FALSE) +
  geom_smooth(linetype = 2) +
  geom_line(aes(SqFtTotLiving, Terms))
```
### Polynomial and Spline Regression
#### Polynomial
```{r paged.print=FALSE}
lm_poly <- lm(AdjSalePrice ~ poly(SqFtTotLiving, 2) + SqFtLot +
    BldgGrade + Bathrooms + Bedrooms, data = house_98105)
lm_poly
```
#### Splines
```{r paged.print=FALSE}
knots <- quantile(house_98105$SqFtTotLiving, p = c(0.25, 0.5, 0.75))
lm_spline <- lm(AdjSalePrice ~ bs(SqFtTotLiving, knots = knots, degree = 3) +
    SqFtLot + Bathrooms + Bedrooms + BldgGrade,
  data = house_98105)
lm_spline
```
#### Generalized Additive Models
```{r paged.print=FALSE}
lm_gam <- gam(AdjSalePrice ~ s(SqFtTotLiving) + SqFtLot +
    Bathrooms + Bedrooms + BldgGrade, data = house_98105)
lm_gam
```
# Supplementary Material
## Figure 4-1. Cotton exposure versus lung capacity
```{r paged.print=FALSE}
lung <- read_csv(file.path(DATA_DIR, "LungDisease.csv"), show_col_types = FALSE)
ggplot(lung, aes(x = Exposure, y = PEFR)) +
  geom_point()
```
## Figure 4-2. Slope and intercept for the regression fit to the lung data
```{r paged.print=FALSE}
model <- lm(PEFR ~ Exposure, data = lung)
x <- c(7.5, 17.75)
y <- predict(model, newdata = data.frame(Exposure = x))

ggplot(lung, aes(x = Exposure, y = PEFR)) +
  geom_abline(intercept = model$coefficients[1], slope = model$coefficients[2],
    color = "#888888") +
  coord_cartesian(xlim = c(0, 25), ylim = c(300, 450), clip = "off") +
  geom_point(x = 0, y = model$coefficients[1], shape = 3, size = 3) +
  annotate("text", -2.25, model$coefficients[1], label = TeX("$b_0$")) +
  geom_segment(x = x[1], y = y[2], xend = x[2], yend = y[2],
    color = "red", linetype = 3) +
  geom_segment(x = x[1], y = y[1], xend = x[1], yend = y[2],
    color = "red", linetype = 3) +
  annotate("text", x = x[1], y = mean(y), label = TeX("$Delta Y$"),
    hjust = 1.5) +
  annotate("text", x = mean(x), y = y[2], label = TeX("$Delta X$"),
    vjust = 2) +
  annotate("text", x = mean(x), y = 390, label = TeX("$b_1 = \\frac{Delta~Y}{Delta~X}$"))
```
## Figure 4-3. Residuals for the regression fit to the lung data
```{r paged.print=FALSE}
fitted <- predict(model)
resid <- residuals(model)

lung1 <- lung %>%
  mutate(
    Fitted = fitted,
    positive = PEFR > Fitted,
  ) %>%
  group_by(Exposure, positive) %>%
  summarize(
    PEFR_max = max(PEFR),
    PEFR_min = min(PEFR),
    Fitted = first(Fitted),
    .groups = "keep",
  ) %>%
  ungroup() %>%
  mutate(PEFR = ifelse(positive, PEFR_max, PEFR_min)) %>%
  arrange(Exposure)

ggplot(lung1, aes(x = Exposure, y = PEFR, xend = Exposure, yend = Fitted)) +
  geom_abline(intercept = model$coefficients[1], slope = model$coefficients[2],
    color = "grey") +
  geom_segment(linetype = 3, color = "red") +
  geom_point()
```
## Figure 4-5. An example of an influential data point in regression
```{r paged.print=FALSE}
seed <- 11
set.seed(seed)

df <- tibble(
  x = rnorm(25),
  y = - x / 5 + rnorm(25)
)
df$x[1] <- 8
df$y[1] <- 8

model1 <- lm(y ~ x, data = df)
model2 <- lm(y[-1] ~ x[-1], data = df)

g <- ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  coord_cartesian(xlim = c(-2.5, 8.5), ylim = c(-2.5, 8.5)) +
  geom_abline(intercept = model1$coefficients[1], slope = model1$coefficients[2],
    col = "blue") +
  geom_abline(intercept = model2$coefficients[1], slope = model2$coefficients[2],
    col = "red", linetype = 2)
g
```
## Figure 4-8. A histogram of the residuals from the regression of the housing data
```{r paged.print=FALSE}
ggplot(tibble(x = df_influential$std_resid), aes(x = x)) +
  geom_histogram(bins = 14) +
  labs(x = "std. residual", y = "Frequency")
```
## Figure 4-10. A polynomial regression fit for the variable SqFtTotLiving
### (solid line) versus a smooth (dashed line; see the following section about splines)
```{r paged.print=FALSE}
terms <- predict(lm_poly, type = "terms")
partial_resid <- resid(lm_poly) + terms
df <- data.frame(SqFtTotLiving = house_98105[, "SqFtTotLiving"],
  Terms = terms[, 1],
  PartialResid = partial_resid[, 1])
graph <- ggplot(df, aes(SqFtTotLiving, PartialResid)) +
  geom_point(shape = 1) +
  scale_shape(solid = FALSE) +
  geom_smooth(linetype = 2, formula = y ~ x, method = "loess") +
  geom_line(aes(SqFtTotLiving, Terms)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
graph
```
## Figure 4-12. A spline regression fit for the variable SqFtTotLiving (solid line) compared to a smooth (dashed line)
```{r paged.print=FALSE}
terms1 <- predict(lm_spline, type = "terms")
partial_resid1 <- resid(lm_spline) + terms1
df1 <- data.frame(SqFtTotLiving = house_98105[, "SqFtTotLiving"],
  Terms = terms1[, 1],
  PartialResid = partial_resid1[, 1])

graph <- ggplot(df1, aes(SqFtTotLiving, PartialResid)) +
  geom_point(shape = 1) +
  scale_shape(solid = FALSE) +
  geom_smooth(linetype = 2, formula = y ~ x, method = "loess") +
  geom_line(aes(SqFtTotLiving, Terms)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
graph
```
## Figure 4-13. A GAM regression fit for the variable SqFtTotLiving (solid line) compared to a smooth (dashed line)
```{r paged.print=FALSE}
terms <- predict.gam(lm_gam, type = "terms")
partial_resid <- resid(lm_gam) + terms
df <- data.frame(SqFtTotLiving = house_98105[, "SqFtTotLiving"],
  Terms = terms[, 5],
  PartialResid = partial_resid[, 5])
graph <- ggplot(df, aes(SqFtTotLiving, PartialResid)) +
  geom_point(shape = 1) +
  scale_shape(solid = FALSE) +
  geom_smooth(linetype = 2, formula = y ~ x, method = "loess") +
  geom_line(aes(SqFtTotLiving, Terms)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
graph
```
## Generalized additive models using statsmodels
## Additional material not covered in book
### Regularization - Lasso
