---
title: "Chapter 6: Statistical Machine Learning"
format: html
---
```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE, fig.align = "center")
```
```{r paged.print=FALSE}
#| cache: FALSE
#| message: FALSE
#| warning: FALSE
library(future)
library(patchwork)
library(ranger)
library(rpart.plot)
library(tidymodels)
library(tidyverse)
library(xgboost)

set.seed(123)

# Location of the data files. Adjust this path if you keep the data
# files in a different directory.
DATA_DIR <- '../../data'

```
## Statistical Machine Learning
### K-Nearest Neighbors
#### A Small Example: Predicting Loan Default
```{r paged.print=FALSE}
loan200 <- read_csv(file.path(DATA_DIR, "loan200.csv"), show_col_types = FALSE) %>%
  mutate(outcome = factor(outcome, levels = c("paid off", "default")))

train_data <- loan200 %>% dplyr::slice(-1)
newloan <- loan200 %>% dplyr::slice(1)
model <- nearest_neighbor(mode = "classification", neighbors = 20,
  weight_func = "rectangular")
knn_model <- workflow() %>%
  add_model(model) %>%
  add_formula(outcome ~ payment_inc_ratio + dti) %>%
  fit(train_data)

knn_pred <- knn_model %>% predict(newloan)
knn_pred
```
#### Standardization (Normalization, z-Scores)
```{r paged.print=FALSE}
newloan
```
```{r paged.print=FALSE}
#| cache: FALSE
library(FNN)
loan_data <- read_csv(file.path(DATA_DIR, "loan_data.csv.gz"), show_col_types = FALSE) %>%
  select(-c(index, status)) %>%
  mutate(
    across(where(is.character), as.factor),
    outcome = factor(outcome, levels = c("paid off", "default"))
  )

loan_df <- model.matrix(~ -1 + payment_inc_ratio + dti + revol_bal +
    revol_util, data = loan_data)
newloan <- loan_df[1, , drop = FALSE]
loan_df <- loan_df[-1, ]
outcome <- loan_data[-1, ]$outcome
knn_pred <- knn(train = loan_df, test = newloan, cl = outcome, k = 5)
loan_df[attr(knn_pred, "nn.index"), ]
```
```{r paged.print=FALSE}
loan_df <- model.matrix(~ -1 + payment_inc_ratio + dti + revol_bal +
    revol_util, data = loan_data)
loan_std <- scale(loan_df)
newloan_std <- loan_std[1, , drop = FALSE]
loan_std <- loan_std[-1, ]
loan_df <- loan_df[-1, ]
outcome <- loan_data[-1, ]$outcome
knn_pred <- knn(train = loan_std, test = newloan_std, cl = outcome, k = 5)
loan_df[attr(knn_pred, "nn.index"), ]
```
#### KNN as a Feature Engine
```{r paged.print=FALSE}
rec <- recipe(outcome ~ dti + revol_bal + revol_util + open_acc +
    delinq_2yrs_zero + pub_rec_zero, data = loan_data)
model <- nearest_neighbor(mode = "classification", neighbors = 20,
  weight_func = "rectangular")
borrow_knn <- workflow() %>%
  add_model(nearest_neighbor(mode = "classification", neighbors = 20)) %>%
  add_recipe(rec) %>%
  fit(loan_data)
borrow_feature <- borrow_knn %>%
  predict(loan_data, type = "prob") %>%
  pluck(".pred_default")
summary(borrow_feature)
```
### Tree Models
#### A Simple Example
```{r paged.print=FALSE}
loan3000 <- read_csv(file.path(DATA_DIR, "loan3000.csv"), show_col_types = FALSE) %>%
  mutate(outcome = factor(outcome, levels = c("paid off", "default")))

loan_tree <- decision_tree(mode = "classification", engine = "rpart") %>%
  fit(outcome ~ borrower_score + payment_inc_ratio, data = loan3000)

rpart.plot(loan_tree %>% extract_fit_engine(), roundint = FALSE)
```
```{r paged.print=FALSE}
loan_tree
```
### Bagging and the Random Forest
#### Random Forest
```{r paged.print=FALSE}
rf <- rand_forest(mode = "classification") %>%
  set_engine("ranger", seed = 123) %>%
  fit(outcome ~ borrower_score + payment_inc_ratio, data = loan3000)
rf
```
```{r paged.print=FALSE}
num_trees <- c(1, seq(5, 500, 5))
oob_scores <- c()
for (trees in num_trees) {
  rf <- rand_forest(mode = "classification", trees = trees) %>%
    set_engine("ranger", seed = 123) %>%
    fit(outcome ~ borrower_score + payment_inc_ratio, data = loan3000)
  oob_scores <- c(oob_scores, (rf %>% extract_fit_engine())$prediction.error)
}
error_df <- tibble(error_rate = oob_scores, num_trees = num_trees)
ggplot(error_df, aes(x = num_trees, y = error_rate)) +
  geom_line()
```
```{r paged.print=FALSE}
rf_df <- augment(rf, loan3000)
ggplot(data = rf_df, aes(x = borrower_score, y = payment_inc_ratio,
    shape = .pred_class, color = .pred_class, size = .pred_class)) +
  geom_jitter(alpha = 0.8, height = 0, width = 0.003) +
  scale_color_manual(values = c("paid off" = "#b8e186", "default" = "#d95f02")) +
  scale_shape_manual(values = c("paid off" = 0, "default" = 1)) +
  scale_size_manual(values = c("paid off" = 0.5, "default" = 2))
```
#### Variable Importance
```{r paged.print=FALSE}
rf_all <- rand_forest(mode = "classification") %>%
  set_engine("ranger", seed = 123, importance = "impurity") %>%
  fit(outcome ~ ., data = loan_data)
rf_all
```
```{r paged.print=FALSE}
rf_permutation <- rand_forest(mode = "classification") %>%
  set_engine("ranger", seed = 123, importance = "permutation") %>%
  fit(outcome ~ ., data = loan_data)
```
```{r paged.print=FALSE}
varimp_gini <- importance(rf_all %>% extract_fit_engine())
varimp_perm <- importance(rf_permutation %>% extract_fit_engine())
df <- tibble(
  predictor = names(varimp_gini),
  impurity = varimp_gini,
  permutation = varimp_perm
)
g1 <- ggplot(df, aes(x = permutation, y = reorder(predictor, permutation))) +
  geom_point() +
  labs(y = "Importance", title = "Accuracy Decrease")
g2 <- ggplot(df, aes(x = impurity, y = reorder(predictor, permutation))) +
  geom_point() +
  labs(y = "Importance", title = "Gini Decrease")
(g1 + theme_bw()) / (g2 + theme_bw())
```
### Boosting
#### XGBoost
```{r paged.print=FALSE}
xgb <- boost_tree(mode = "classification",
  trees = 100, learn_rate = 0.1, sample_size = 0.63) %>%
  fit(outcome ~ borrower_score + payment_inc_ratio, data = loan3000)
```
```{r paged.print=FALSE}
xgb_df <- augment(xgb, loan3000)
ggplot(data = xgb_df, aes(x = borrower_score, y = payment_inc_ratio,
    color = .pred_class, shape = .pred_class, size = .pred_class)) +
  geom_jitter(alpha = 0.8, height = 0, width = 0.003) +
  scale_color_manual(values = c("paid off" = "#b8e186", "default" = "#d95f02")) +
  scale_shape_manual(values = c("paid off" = 0, "default" = 1)) +
  scale_size_manual(values = c("paid off" = 0.5, "default" = 2))
```
#### Regularization: Avoiding Overfitting
```{r paged.print=FALSE}
set.seed(400820)
loan_split <- initial_split(loan_data, prop = 0.75)
train_data <- training(loan_split)
test_data <- testing(loan_split)
train_x <- data.matrix(train_data %>% select(-outcome))
train_y <- factor(train_data$outcome, levels = c("paid off", "default"))
test_x <- data.matrix(test_data %>% select(-outcome))
test_y <- factor(test_data$outcome, levels = c("paid off", "default"))

xgb_default <- xgboost(x = train_x, y = train_y,
  objective = "binary:logistic", nrounds = 250, verbosity = 1,
  eval_metric = "error")

pred_default <- predict(xgb_default, test_x)
error_default <- abs((as.numeric(test_y) - 1) - pred_default) > 0.5
attributes(xgb_default)$evaluation_log[250, ]
mean(error_default)
```
```{r paged.print=FALSE}
xgb_penalty <- xgboost(x = train_x, y = train_y,
  learning_rate = 0.1, subsample = 0.63, reg_lambda = 1000,
  objective = "binary:logistic", nrounds = 250, verbosity = 1,
  eval_metric = "error")
pred_penalty <- predict(xgb_penalty, test_x)
error_penalty <- abs((as.numeric(test_y) - 1) - pred_penalty) > 0.5
attributes(xgb_penalty)$evaluation_log[250, ]
mean(error_penalty)
```
```{r paged.print=FALSE}
error_default <- rep(0, 249)
error_penalty <- rep(0, 249)
for (ntree_limit in 1:249) {
  iteration_range <- c(1, ntree_limit + 1)
  pred_def <- predict(xgb_default, test_x, iteration_range = iteration_range)
  error_default[ntree_limit] <- mean(abs(as.numeric(test_y)-1-pred_def) >= 0.5)
  pred_pen <- predict(xgb_penalty, test_x, iteration_range = iteration_range)
  error_penalty[ntree_limit] <- mean(abs(as.numeric(test_y)-1-pred_pen) >= 0.5)
}
```
```{r paged.print=FALSE}
#| cache: FALSE
errors <- bind_rows(
  tibble(Iterations = 1:250, Dataset = "train", Penalty = "no",
    Error = attributes(xgb_default)$evaluation_log$train_error),
  tibble(Iterations = 1:249, Dataset = "test", Penalty = "no",
    Error = error_default),
  tibble(Iterations = 1:250, Dataset = "train", Penalty = "yes",
    Error = attributes(xgb_penalty)$evaluation_log$train_error),
  tibble(Iterations = 1:249, Dataset = "test", Penalty = "yes",
    Error = error_penalty),
)
ggplot(errors, aes(x = Iterations, y = Error, color = Penalty,
    linetype = Dataset)) +
  geom_line() +
  scale_color_manual(values = c("red", "black"))
```
```{r paged.print=FALSE}
#| cache: FALSE
last_plot()  +
  theme_bw() +
  annotate("text", 165, 0.2, label="Default / Training set") +
  annotate("text", 150, 0.3, label="Penalty / Training set") +
  annotate("text", 220, 0.34, label="Penalty / Test set") +
  annotate("text", 75, 0.36, label="Default / Test set") +
  theme(legend.position = "none")
```
```{r paged.print=FALSE}
#| cache: FALSE
```
#### Hyperparameters and Cross-Validation
```{r paged.print=FALSE}
set.seed(1234)
resamples <- vfold_cv(loan_data, v = 5)
grid <- crossing(
  tree_depth = c(3, 6, 12),
  learn_rate = c(0.1, 0.5, 0.9),
)
```
```{r paged.print=FALSE}
plan(multisession, workers = parallel::detectCores(logical = FALSE))

rec <- recipe(outcome ~ ., data = loan_data) %>%
  step_dummy(all_nominal_predictors())
model <- boost_tree(mode = "classification",
  tree_depth = tune(), learn_rate = tune())

wf <- workflow() %>%
  add_model(model) %>%
  add_recipe(rec)

model_cv <- tune_grid(wf, resamples, grid = grid)
```
```{r paged.print=FALSE}
collect_metrics(model_cv) %>%
  filter(.metric == "accuracy") %>%
  mutate(error_rate = round(100 * (1 - mean), 2)) %>%
  select(c(tree_depth, learn_rate, error_rate)) %>%
  arrange(error_rate)
```
# Supplementary Material
## Figure 6-2. KNN prediction of loan default using two variables: debt-to-income ratio and loan-payment-to-income ratio
```{r paged.print=FALSE}
train_data <- loan200 %>%
  dplyr::slice(-1)
newloan <- loan200 %>%
  dplyr::slice(1) %>%
  select(payment_inc_ratio, dti)

# The knn function from the class package gives us access to a list of the nearest
# neighbors and their distances via attributes "nn.index" and "nn.dist"
knn_pred <- knn(train = train_data[c("payment_inc_ratio", "dti")], test = newloan,
  cl = train_data$outcome, k = 20)

nearest_points <- loan200[c(attr(knn_pred, "nn.index") + 1), ]
nearest_points
dist <- attr(knn_pred, "nn.dist")

circle_coordinates <- function(center = c(0, 0), r = 1, npoints = 100) {
  tt <- seq(0, 2 * pi, length.out = npoints - 1)
  xx <- center[1] + r * cos(tt)
  yy <- center[2] + r * sin(tt)
  return(data.frame(x = c(xx, xx[1]), y = c(yy, yy[1])))
}

circle_df <- circle_coordinates(center = unlist(newloan), r = max(dist),
  npoints = 201)
```
```{r paged.print=FALSE}
# set first entry as target - requires adding additional level to factor
levels(loan200$outcome) <- c(levels(loan200$outcome), "newloan")
loan200[1, "outcome"] <- "newloan"
head(loan200)
levels(nearest_points$outcome)
```
```{r paged.print=FALSE}
graph <- ggplot(data = loan200, aes(x = payment_inc_ratio, y = dti, color = outcome)) +
  geom_point(aes(shape = outcome), size = 2, alpha = 0.4) +
  geom_point(data = nearest_points, aes(shape = outcome), size = 2) +
  geom_point(data = loan200[1, ], aes(shape = outcome), size = 2) +
  scale_shape_manual(values = c(15, 16, 4)) +
  scale_color_manual(values = c("paid off" = "#1b9e77", "default" = "#d95f02", "newloan" = "black")) +
  geom_path(data = circle_df, aes(x = x, y = y), color = "black") +
  coord_cartesian(xlim = c(3, 15), ylim = c(17, 29)) +
  theme_bw()
graph
```
## Figure 6-4. The first five rules for a simple tree model fit to the loan data
```{r paged.print=FALSE}
## Figure 6-4: View of partition rules
r_tree <- tibble(
  x1 = c(0.575, 0.375, 0.375, 0.375, 0.475),
  x2 = c(0.575, 0.375, 0.575, 0.575, 0.475),
  y1 = c(0, 0, 10.42, 4.426, 4.426),
  y2 = c(25, 25, 10.42, 4.426, 10.42),
  rule_number = factor(c(1, 2, 3, 4, 5)))
r_tree <- as.data.frame(r_tree)

rules <- tibble(
  x = c(0.575, 0.375, 0.4, 0.4, 0.475),
  y = c(24, 24, 10.42, 4.426, 9.42),
  rule_number = factor(c(1, 2, 3, 4, 5))
)

labs <- tibble(
  x = c(
    0.575 + (1 - 0.575) / 2,
    0.375 / 2,
    (0.375 + 0.575) / 2,
    (0.375 + 0.575) / 2,
    (0.475 + 0.575) / 2,
    (0.375 + 0.475) / 2
  ),
  y = c(
    12.5,
    12.5,
    10.42 + (25 - 10.42) / 2,
    4.426 / 2,
    4.426 + (10.42 - 4.426) / 2,
    4.426 + (10.42 - 4.426) / 2
  ),
  decision = factor(c(
    "paid off",
    "default",
    "default",
    "paid off",
    "paid off",
    "default"
  ))
)
```
```{r paged.print=FALSE}
shift <- 0.004
shift <- 0
graph <- loan3000 %>%
  #  mutate(borrower_score = ifelse(outcome == "paid off", borrower_score + shift, borrower_score - shift)) %>%
  mutate(
    borrower_score = borrower_score + rnorm(dim(loan3000)[1], sd = 2 * shift)
  ) %>%
  ggplot(aes(x = borrower_score, y = payment_inc_ratio)) +
    geom_point(aes(color = outcome, size = outcome, shape = outcome), alpha = 0.5) +
    # scale_color_manual(values = c("paid off" = "#7fbc41", "default" = "#d95f02")) +
    scale_shape_manual(values = c("paid off" = 0, "default" = 1)) +
    scale_size_manual(values = c("paid off" = 1, "default" = 1)) +
    geom_segment(data = r_tree, aes(x = x1, y = y1, xend = x2, yend = y2), linewidth = 1.5) +
    guides(color = guide_legend(override.aes = list(linewidth = 1.5)),
      linetype = guide_legend(keywidth = 3, override.aes = list(linewidth = 1))) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    coord_cartesian(ylim = c(0, 25)) +
    geom_label(data = labs, aes(x = x, y = y, label = decision)) +
    geom_label(data = rules, aes(x = x, y = y, label = rule_number),
      size = 2.5, fill = "#eeeeee",
      label.r = unit(0, "lines"), label.padding = unit(0.2, "lines")) +
    guides(color = guide_legend(override.aes = list(size = 2))) +
    stat_density_2d(aes(linetype = outcome, color = outcome),  
    # variation geom = "polygon",
    # variation aes(alpha = 0.1 * ..level.., fill = outcome),
                    bins = 6)  +
    theme_bw()
graph
```
```{r paged.print=FALSE}
graph <- ggplot(data = loan3000, aes(x = borrower_score, y = payment_inc_ratio)) +
  geom_point(aes(color = outcome, shape = outcome, size = outcome), alpha = 0.8) +
  scale_color_manual(values = c("paid off" = "#7fbc41", "default" = "#d95f02")) +
  scale_shape_manual(values = c("paid off" = 0, "default" = 1)) +
  scale_size_manual(values = c("paid off" = 0.5, "default" = 2)) +
  geom_segment(data = r_tree, aes(x = x1, y = y1, xend = x2, yend = y2), size = 1.5) +
  guides(color = guide_legend(override.aes = list(size = 1.5)),
    linetype = guide_legend(keywidth = 3, override.aes = list(size = 1))) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim = c(0, 25)) +
  geom_label(data = labs, aes(x = x, y = y, label = decision)) +
  geom_label(data = rules, aes(x = x, y = y, label = rule_number),
    size = 2.5, fill = "#eeeeee",
    label.r = unit(0, "lines"), label.padding = unit(0.2, "lines")) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_bw()
graph
```
## Figure 6-5. Gini impurity and entropy measures
```{r paged.print=FALSE}
info <- function(x) {
  info <- ifelse(x == 0, 0, -x * log2(x) - (1 - x) * log2(1 - x))
  return(info)
}

gini <- function(x) {
  return(x * (1 - x))
}

x <- 0:50 / 100
impure <- bind_rows(
  tibble(p = x, impurity = 2 * x, type = "Accuracy"),
  tibble(p = x, impurity = gini(x) / gini(0.5) * info(0.5), type = "Gini"),
  tibble(p = x, impurity = info(x), type = "Entropy")
)

graph <- ggplot(data = impure, aes(x = p, y = impurity, linetype = type, color = type)) +
  geom_line(size = 1.5) +
  guides(linetype = guide_legend(keywidth = 3, override.aes = list(size = 1))) +
  scale_x_continuous(expand = c(0, 0.01)) +
  scale_y_continuous(expand = c(0, 0.01)) +
  theme_bw() +
  theme(legend.title = element_blank())
graph
```
