---
title: "Chapter 2: Data and Sampling Distributions"
format: html
---
```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE, fig.align = "center")
```

```{r paged.print=FALSE}
#| cache: FALSE
#| message: FALSE
#| warning: FALSE
library(boot)
library(tidyverse)

set.seed(123)

# Location of the data files. Adjust this path if you keep the data
# files in a different directory.
DATA_DIR <- '../data'

```
## Data and Sampling Distributions
### Sampling Distribution of a Statistic
```{r paged.print=FALSE}
loans_income <- read_csv(file.path(DATA_DIR, "loans_income.csv"), show_col_types = FALSE) %>%
  pull(x)

# take a simple random sample
samp_data <- data.frame(income = sample(loans_income, 1000),
  type = "Data")
# take a sample of means of 5 values
samp_mean_05 <- tibble(
  income = replicate(1000, mean(sample(loans_income, 5))),
  type = "Mean of 5"
)
# take a sample of means of 20 values
samp_mean_20 <- tibble(
  income = replicate(1000, mean(sample(loans_income, 20))),
  type = "Mean of 20"
)
income <- bind_rows(samp_data, samp_mean_05, samp_mean_20) %>%
  mutate(type = factor(type,
      levels = c("Data", "Mean of 5", "Mean of 20"))
  )

# plot the histograms
ggplot(income, aes(x = income)) +
  geom_histogram(bins = 40) +
  facet_grid(type ~ .)
```
### The Bootstrap
```{r paged.print=FALSE}
stat_fun <- function(x, idx) median(x[idx])
boot_obj <- boot(loans_income, R = 1000, statistic = stat_fun)
```
```{r paged.print=FALSE}
boot_obj
```
### Normal Distribution
#### Standard Normal and QQ-Plots
```{r paged.print=FALSE}
norm_samp <- rnorm(100)
ggplot(tibble(y = norm_samp), aes(sample = y)) +
  geom_qq() +
  geom_qq_line() +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles")
```
```{r paged.print=FALSE}
# alternative using qqnorm
qqnorm(norm_samp)
qqline(norm_samp)
```
### Long-Tailed Distributions
```{r paged.print=FALSE}
sp500_px <- read_csv(file.path(DATA_DIR, "sp500_data.csv.gz"), show_col_types = FALSE)

nflx <- sp500_px[, "NFLX"]
nflx <- diff(log(nflx[nflx > 0]))
qqnorm(nflx)
abline(a = 0, b = 1, col = "grey")
ggplot(tibble(y = nflx), aes(sample = y)) +
  geom_qq() +
  geom_qq_line() +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles")
```
### Binomial Distribution
```{r paged.print=FALSE}
dbinom(x = 2, size = 5, prob = 0.1)
```
```{r paged.print=FALSE}
pbinom(q = 2, size = 5, prob = 0.1)
```
### Poisson and Related Distributions
#### Poisson Distributions
```{r paged.print=FALSE}
rpois(n = 100, lambda = 2)
```
#### Exponential Distribution
```{r paged.print=FALSE}
rexp(n = 100, rate = 0.2)
```
#### Weibull Distribution
```{r paged.print=FALSE}
rweibull(n = 100, shape = 1.5, scale = 5000)
```
# Supplementary Material
## Figure 2-9 Bootstrap confidence interval for the annual income of loan applicants, based on a sample of 20
```{r paged.print=FALSE}
loans_income <- read_csv(file.path(DATA_DIR, "loans_income.csv"), show_col_types = FALSE) %>%
  pull(x)

set.seed(7)
sample20 <- sample(loans_income, 20)
sample_mean <- mean(sample20)

stat_fun <- function(x, idx) mean(x[idx])
boot_obj <- boot(sample20, R = 500, statistic = stat_fun)
boot_ci <- boot.ci(boot_obj, conf = 0.9, type = "basic")
df <- data.frame(mean = boot_obj$t)
ci90 <- boot_ci$basic[4:5]
ci <- data.frame(ci = ci90, y = c(9, 11))
ci

ggplot(df, aes(x = mean)) +
  geom_histogram(bins = 40, fill = "#AAAAAA") +
  geom_vline(xintercept = sample_mean, linetype = 2) +
  geom_path(aes(x = ci, y = 10), data = ci, linewidth = 2) +
  geom_path(aes(x = ci90[1], y = y), data = ci, linewidth = 2) +
  geom_path(aes(x = ci90[2], y = y), data = ci, linewidth = 2) +
  annotate("text", x = sample_mean, y = 20, label = "Sample mean", size = 6) +
  annotate("text", x = sample_mean, y = 8, label = "90% interval", size = 6) +
  theme_bw() +
  labs(x = "", y = "Counts")
```
## Visualization of Distributions
### Poisson Distribution
### Exponential Distribution
### Weibull Distribution
