---
title: "Chapter 5: Classification"
format: html
---
```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE, fig.align = "center")
```
```{r paged.print=FALSE}
#| cache: FALSE
#| message: FALSE
#| warning: FALSE
library(discrim)
library(ggpubr)
library(mgcv)
library(patchwork)
library(rpart)
library(tidymodels)
library(tidyverse)

set.seed(123)

# Location of the data files. Adjust this path if you keep the data
# files in a different directory.
DATA_DIR <- '../data'

```
## Classification
### Naive Bayes
#### The Naive Solution
```{r paged.print=FALSE}
loan_data <- read_csv(file.path(DATA_DIR, "loan_data.csv.gz"), show_col_types = FALSE) %>%
  dplyr::select(-c(index, status)) %>%
  mutate(
    across(where(is.character), as.factor),
    outcome = factor(outcome, levels = c("paid off", "default")),
  )
naive_model <- naive_Bayes() %>%
  fit(outcome ~ purpose_ + home_ + emp_len_, loan_data)

naive_model %>%
  extract_fit_engine() %>%
  pluck("tables")
```
```{r paged.print=FALSE}
new_loan <- loan_data[147, c("purpose_", "home_", "emp_len_")]
row.names(new_loan) <- NULL
new_loan
```
```{r paged.print=FALSE}
predict(naive_model, new_loan)
```
```{r paged.print=FALSE}
predict(naive_model, new_loan, type = "prob")
```
#### Numeric Predictor Variables
```{r paged.print=FALSE}
less_naive <- naive_Bayes() %>%
  set_args(usekernel = FALSE) %>%
  fit(outcome ~ borrower_score + payment_inc_ratio +
      purpose_ + home_ + emp_len_, loan_data)

(naive_model %>%
    extract_fit_engine() %>%
    pluck("tables"))[1:2]
```
```{r paged.print=FALSE}
p_paid <- sum(loan_data$outcome == "default") / nrow(loan_data)
p_default <- sum(loan_data$outcome == "paid off") / nrow(loan_data)
p_bs_paid <- function(x, stats) dnorm(x, mean = stats[1, 1], sd = stats[1, 2])
p_bs_default <- function(x, stats) dnorm(x, mean = stats[2, 1], sd = stats[2, 2])
p_paid_bs <- function(x, stats) {
  return(p_paid * p_bs_paid(x, stats) /
      (p_paid * p_bs_paid(x, stats) + p_default * p_bs_default(x, stats)))
}
p_default_bs <- function(x, stats) {
  return(p_default * p_bs_default(x, stats) /
      (p_paid * p_bs_paid(x, stats) + p_default * p_bs_default(x, stats)))
}

stats <- less_naive %>%
  extract_fit_engine() %>%
  pluck("tables", "borrower_score")
g1 <- ggplot(data.frame(borrower_score = c(0, 1)), aes(x = borrower_score)) +
  stat_function(fun = p_bs_paid, color = "blue", linetype = 1,
    args = list(stats = stats)) +
  stat_function(fun = p_bs_default, color = "red", linetype = 2,
    args = list(stats = stats)) +
  labs(y = "P(Xj | Y=i)") + theme_bw()
g2 <- ggplot(data.frame(borrower_score = c(0, 1)), aes(x = borrower_score)) +
  stat_function(fun = p_paid_bs, color = "blue", linetype = 1,
    args = list(stats = stats)) +
  stat_function(fun = p_default_bs, color = "red", linetype = 2,
    args = list(stats = stats)) +
  labs(y = "P(Y=i | Xj)") + theme_bw()
g <- wrap_plots(g1, g2)
g
```
### Discriminant Analysis
#### A Simple Example
```{r paged.print=FALSE}
loan3000 <- read_csv(file.path(DATA_DIR, "loan3000.csv"), show_col_types = FALSE) %>%
  mutate(
    outcome = factor(outcome, levels = c("paid off", "default")),
    purpose_ = factor(purpose_)
  )

loan_lda <- discrim_linear() %>%
  fit(outcome ~ borrower_score + payment_inc_ratio, data = loan3000)

loan_lda %>%
  extract_fit_engine() %>%
  pluck("scaling")
```
```{r paged.print=FALSE}
lda_pred <- augment(loan_lda, new_data = loan3000)
head(lda_pred %>% select(.pred_class, .pred_default, `.pred_paid off`))
```
```{r paged.print=FALSE}
loan_fit_engine <- loan_lda %>% extract_fit_engine()
center <- 0.5 * (loan_fit_engine$mean[1, ] + loan_fit_engine$mean[2, ])
slope <- -loan_fit_engine$scaling[1] / loan_fit_engine$scaling[2]
intercept <- center[2] - center[1] * slope

ggplot(data = lda_pred,
  aes(x = borrower_score, y = payment_inc_ratio, color = .pred_default)) +
  geom_point(alpha = 0.6) +
  scale_color_gradientn(colors = c("#ca0020", "#f7f7f7", "#0571b0")) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 20)) +
  geom_abline(slope = slope, intercept = intercept, color = "darkgreen")
```
### Logistic Regression
#### Logistic Regression and the GLM
```{r paged.print=FALSE}
logistic_model <- logistic_reg(mode = "classification") %>%
  fit(outcome ~ payment_inc_ratio + purpose_ +
      home_ + emp_len_ + borrower_score, loan_data)
logistic_model
```
#### Predicted Values from Logistic Regression
```{r paged.print=FALSE}
pred <- predict(logistic_model, new_data = loan_data, type = "raw")
summary(pred)
```
```{r paged.print=FALSE}
prob <- 1 / (1 + exp(-pred))
summary(prob)
```
```{r paged.print=FALSE}
prob <- predict(logistic_model, new_data = loan_data, type = "prob")
summary(prob$.pred_default)
```
#### Assessing the Model
```{r paged.print=FALSE}
logistic_model %>%
  extract_fit_engine() %>%
  summary()
```
```{r paged.print=FALSE}
gam_formula <- outcome ~ s(payment_inc_ratio) + purpose_ +
  home_ + emp_len_ + s(borrower_score)
logistic_gam <- gen_additive_mod(mode = "classification") %>%
  fit(gam_formula, data = loan_data)
```
##### Analysis of residuals
```{r paged.print=FALSE}
terms <- predict(logistic_gam %>% extract_fit_engine(), type = "terms")
partial_resid <- resid(logistic_gam %>% extract_fit_engine()) + terms
df <- data.frame(payment_inc_ratio = loan_data[, "payment_inc_ratio"],
  terms = terms[, "s(payment_inc_ratio)"],
  partial_resid = partial_resid[, "s(payment_inc_ratio)"])
ggplot(df, aes(x = payment_inc_ratio, y = partial_resid, solid = FALSE)) +
  geom_point(shape = 46, alpha = 0.4) +
  geom_line(aes(x = payment_inc_ratio, y = terms),
    color = "red", alpha = 0.5, linewidth = 1.5) +
  labs(y = "Partial Residual") +
  coord_cartesian(xlim = c(0, 25))
```
### Evaluating Classification Models
####  Confusion Matrix
```{r paged.print=FALSE}
predictions <- augment(logistic_gam, loan_data)
pred_y <- as.numeric(predictions$.pred_class == "default")
true_y <- as.numeric(predictions$outcome == "default")
true_pos <- (true_y == 1) & (pred_y == 1)
true_neg <- (true_y == 0) & (pred_y == 0)
false_pos <- (true_y == 0) & (pred_y == 1)
false_neg <- (true_y == 1) & (pred_y == 0)
conf_mat <- matrix(
  c(sum(true_neg), sum(false_pos),
    sum(false_neg), sum(true_pos)), 2, 2)
rownames(conf_mat) <- c("Yhat = 0", "Yhat = 1")
colnames(conf_mat) <- c("Y = 0", "Y = 1")
conf_mat
```
```{r paged.print=FALSE}
conf_mat(predictions, truth = outcome, estimate = .pred_class)
```
#### Precision, Recall, and Specificity
```{r paged.print=FALSE}
bind_rows(
  precision(predictions, truth = outcome, estimate = .pred_class),
  recall(predictions, truth = outcome, estimate = .pred_class),
  specificity(predictions, truth = outcome, estimate = .pred_class),
) %>% select(-.estimator)
```
#### ROC Curve
```{r paged.print=FALSE}
roc_curve(predictions, outcome, .pred_default, event_level = "second") %>%
  autoplot()
```
```{r paged.print=FALSE}
prob <- predict(logistic_gam, new_data = loan_data, type = "prob")
idx <- order(-prob$.pred_default)
recall <- cumsum(true_y[idx] == 1) / sum(true_y == 1)
specificity <- (sum(true_y == 0) - cumsum(true_y[idx] == 0)) / sum(true_y == 0)
roc_df <- data.frame(recall = recall, specificity = specificity)
ggplot(roc_df, aes(x = specificity, y = recall)) +
  geom_line(color = "blue") +
  scale_x_reverse(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_line(data = data.frame(x = (0:100) / 100), aes(x = x, y = 1 - x),
    linetype = "dotted", color = "red")
```
#### AUC
```{r paged.print=FALSE}
sum(roc_df$recall[-1] * diff(1 - roc_df$specificity))
```
### Strategies for Imbalanced Data
#### Undersampling
```{r paged.print=FALSE}
full_train_set <- read_csv(file.path(DATA_DIR, "full_train_set.csv.gz"), show_col_types = FALSE) %>%
  mutate(
    across(where(is.character), as.factor),
    outcome = factor(outcome, levels = c("paid off", "default")),
  )

mean(full_train_set$outcome == "default")
```
```{r paged.print=FALSE}
full_model <- logistic_reg(mode = "classification") %>%
  fit(outcome ~ payment_inc_ratio + purpose_ +
      home_ + emp_len_ + dti + revol_bal + revol_util,
    full_train_set)
pred <- predict(full_model, new_data = full_train_set)
100 * mean(pred == "default")
```
#### Oversampling and Up/Down Weighting
```{r paged.print=FALSE}
full_train_set <- full_train_set %>%
  mutate(
    weight = ifelse(outcome == "default", 1 / mean(outcome == "default"), 1.0),
    weight = importance_weights(weight))
weighted_model <- workflow() %>%
  add_model(logistic_reg(mode = "classification")) %>%
  add_case_weights(weight) %>%
  add_formula(outcome ~ payment_inc_ratio + purpose_ +
      home_ + emp_len_ + dti + revol_bal + revol_util) %>%
  fit(full_train_set)
pred <- predict(weighted_model, new_data = full_train_set)
100 * mean(pred == "default")
```
# Supplementary Material
## Figure 5-2. Graph of the logit function that maps a probability to a scale suitable for a linear model
```{r paged.print=FALSE}
p <- seq(from = 0.01, to = 0.99, by = 0.01)
df <- tibble(
  p = p,
  logit = log(p / (1 - p)),
  odds = p / (1 - p),
)

graph <- ggplot(data = df, aes(x = p, y = logit)) +
  geom_line() +
  labs(x = "p", y = "logit(p)")
graph
```
## How to control the order of the classes in Python
## Figure 5-3. The relationship between the odds ratio and the log-odds ratio
```{r paged.print=FALSE}
graph <- ggplot(data = df, aes(x = logit, y = odds)) +
  geom_line() +
  labs(x = "log(odds ratio)", y = "odds ratio") +
  coord_cartesian(xlim = c(0, 5), ylim = c(1, 100))
graph
```
## Figure 5-4. Partial residuals from logistic regression
## Figure 5-7. Area under the ROC curve for the loan data
```{r paged.print=FALSE}
graph <- ggplot(roc_df, aes(specificity)) +
  geom_ribbon(aes(ymin = 0, ymax = recall), fill = "blue", alpha = 0.3) +
  scale_x_reverse(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(y = "recall") +
  theme(plot.margin = unit(c(5.5, 10, 5.5, 5.5), "points"))
graph
```
## SMOTE
## Figure 5-8. Comparison of the classification rules for four different methods
```{r paged.print=FALSE}
loan_tree <- rpart(outcome ~ borrower_score + payment_inc_ratio,
  data = loan3000, control = rpart.control(cp = 0.005))
loan_tree

lda_pred <- tibble(
  borrower_score = c((0 - intercept) / slope, (20 - intercept) / slope),
  payment_inc_ratio = c(0, 20),
  method = "LDA",
)

tree_pred <- tibble(
  borrower_score = c(0.375, 0.375, 0.475, 0.475, 0.575, 0.575),
  payment_inc_ratio = c(0, 4.426,  4.426, 10.42, 10.42, 20),
  method = "Tree",
)

glm0 <- glm(outcome ~ payment_inc_ratio +  borrower_score,
  data = loan3000, family = "binomial")
y <- seq(from = 0, to = 20, length = 100)
x <- (- glm0$coefficients[1] - glm0$coefficients[2] * y) / glm0$coefficients[3]
glm0_pred <- tibble(
  borrower_score = x,
  payment_inc_ratio = y,
  method = "Logistic",
)

gam1 <- gam(outcome ~ s(payment_inc_ratio) +  s(borrower_score),
  data = loan3000, family = "binomial")
gam_fun <- function(x) {
  newdata <- data.frame(borrower_score = x, payment_inc_ratio = y)
  rss <- sum(predict(gam1, newdata = newdata)^2)
  return(rss)
}
est_x <- nlminb(seq(from = 0.33, to = 0.73, length = 100), gam_fun)
gam1_pred <- tibble(
  borrower_score = est_x$par,
  payment_inc_ratio = y,
  method = "GAM",
)

loan_fits <- bind_rows(
  lda_pred,
  tree_pred,
  glm0_pred,
  gam1_pred,
)
```
```{r paged.print=FALSE}
graph <- ggplot(data = loan_fits, aes(x = borrower_score, y = payment_inc_ratio, color = method, linetype = method)) +
  geom_line(linewidth = 1.2) +
  theme_bw() +
  theme(legend.key.width = unit(3, "cm")) +
  guides(linetype = guide_legend(override.aes = list(size = 1)))
graph
```
# Naive Bayes with numerical features
# print(f"MixedNB Library Accuracy: {accuracy_mixed:.2f}")
