---
title: "Chapter 7: Unsupervised Learning"
format: html
---
```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE, fig.align = "center")
```
```{r paged.print=FALSE}
#| cache: FALSE
#| message: FALSE
#| warning: FALSE
library(ca)
library(cluster)
library(ellipse)
library(ggrepel)
library(mclust)
library(tidyclust)
library(tidymodels)
library(tidyverse)

set.seed(123)

# Location of the data files. Adjust this path if you keep the data
# files in a different directory.
DATA_DIR <- '../data'

```
## Unsupervised Learning
### Principal Components Analysis
#### A Simple Example
```{r paged.print=FALSE}
sp500_px <- read_csv(file.path(DATA_DIR, "sp500_data.csv.gz"), show_col_types = FALSE)

oil_px <- sp500_px[, c("CVX", "XOM")]
pca <- princomp(oil_px)
pca$loadings
```
```{r paged.print=FALSE}
pca_rec <- recipe(data = sp500_px, formula = ~ CVX + XOM) %>%
  step_pca(all_numeric_predictors())
pca_prep <- pca_rec %>%
  prep()
pca_prep$steps[[1]]$rotation
```
```{r paged.print=FALSE}
loadings <- pca$loadings
ggplot(data = oil_px, aes(x = CVX, y = XOM)) +
  geom_point(alpha = 0.3) +
  stat_ellipse(type = "norm", level = 0.99) +
  geom_abline(intercept = 0, slope = loadings[2, 1] / loadings[1, 1]) +
  geom_abline(intercept = 0, slope = loadings[2, 2] / loadings[1, 2]) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3))
```
#### Interpreting Principal Components
```{r paged.print=FALSE}
syms <- c("AAPL", "MSFT", "CSCO", "INTC", "CVX", "XOM", "SLB",
  "COP", "JPM", "WFC", "USB", "AXP", "WMT", "TGT", "HD", "COST")
top_sp <- sp500_px %>%
  filter(Date >= "2011-01-01") %>%
  select(all_of(syms))
sp_pca <- princomp(top_sp)
screeplot(sp_pca)
```
```{r paged.print=FALSE}
components <- 1:length(sp_pca$sdev)
g <- tibble(
  Components = components,
  Variances = sp_pca$sdev ** 2,
) |>
filter(
  Components < 11,
) |>
  ggplot(aes(x=Components, y=Variances)) +
    geom_bar(stat="identity") +
    coord_cartesian(xlim = c(0.5, 10.49)) +
    scale_x_continuous(breaks = components) +
    theme_bw()
g
```
```{r paged.print=FALSE}
loadings <- sp_pca$loadings[, 1:5]
loadings <- as_tibble(loadings) %>%
  mutate(Symbol = row.names(loadings)) %>%
  pivot_longer(cols = starts_with("Comp"), names_to = "Component",
    values_to = "Weight") %>%
  mutate(color = Weight > 0)
ggplot(loadings, aes(x = Symbol, y = Weight, fill = color)) +
  geom_bar(stat = "identity") +
  facet_grid(Component ~ ., scales = "free_y") +
  guides(fill = "none") +
  labs(y = "Component Loading") +
  theme_bw() +
  theme(axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5))
```
#### Correspondence Analysis
```{r paged.print=FALSE}
housetasks <- read_csv(file.path(DATA_DIR, "housetasks.csv"), show_col_types = FALSE) %>%
  column_to_rownames(var = "Task")
ca_analysis <- ca(housetasks)
plot(ca_analysis)
```
### K-Means Clustering
#### A Simple Example
```{r paged.print=FALSE}
df <- sp500_px %>%
  filter(Date >= "2011-01-01")
kmeans_wf <- workflow() %>%
  add_recipe(recipe(~ XOM + CVX, data = df)) %>%
  add_model(k_means(num_clusters = 4))
model <- kmeans_wf %>% fit(df)
```
```{r paged.print=FALSE}
augment(model, df) %>%
  select(Date, XOM, CVX, .pred_cluster) %>%
  head()
```
```{r paged.print=FALSE}
tidy(model)
```
```{r paged.print=FALSE}
ggplot(augment(model, df), aes(x = XOM, y = CVX)) +
  geom_point(aes(color = .pred_cluster, shape = .pred_cluster), alpha = 0.3) +
  geom_point(data = tidy(model), aes(x = XOM, y = CVX), size = 3, stroke = 2)
```
#### K-Means Algorithm
```{r paged.print=FALSE}
syms <- c("AAPL", "MSFT", "CSCO", "INTC", "CVX", "XOM", "SLB", "COP",
  "JPM", "WFC", "USB", "AXP", "WMT", "TGT", "HD", "COST")
df <- sp500_px %>%
  filter(Date >= "2011-01-01") %>%
  select(all_of(syms))
set.seed(10010)
kmeans_wf <- workflow() %>%
  add_recipe(recipe(~ ., data = df)) %>%
  add_model(k_means(num_clusters = 5) %>%
      set_engine("stats", nstart = 10))
km <- kmeans_wf %>% fit(df)
```
#### Interpreting the Clusters
```{r paged.print=FALSE}
tidy(km)$size
```
```{r paged.print=FALSE}
centers <- tidy(km) %>%
  pivot_longer(cols = all_of(syms), names_to = "Symbol",
    values_to = "Mean") %>%
  mutate(Color = Mean > 0)
ggplot(centers, aes(x = Symbol, y = Mean, fill = Color)) +
  geom_bar(stat = "identity", position = "identity", width = 0.75) +
  facet_grid(cluster ~ ., scales = "free_y") +
  guides(fill = "none") +
  labs(y = "Component Loading") +
  theme_bw() +
  theme(axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5))
```
#### Selecting the Number of Clusters
```{r paged.print=FALSE}
set.seed(10010)
train_kmeans <- function(num_clusters, data) {
  model <- workflow() %>%
    add_recipe(recipe(~ ., data = data)) %>%
    add_model(k_means(num_clusters = num_clusters) %>%
        set_engine("stats", nstart = 50, iter.max = 100)) %>%
    fit(data)
  return(model)
}
km <- train_kmeans(14, df)
totalss <- extract_fit_engine(km)$totss

pct_var <- tibble(pct_var = 0, num_clusters = 2:14)
for (num_clusters in 2:14) {
  km_cluster <- train_kmeans(num_clusters, df)
  pct_var[num_clusters - 1, "pct_var"] <- (
    extract_fit_engine(km_cluster)$betweenss / totalss)
}
ggplot(pct_var, aes(x = num_clusters, y = pct_var)) +
  geom_line() +
  geom_point() +
  labs(x = "Number of clusters (k)", y = "% Variance Explained")
```
### Hierarchical Clustering
#### A Simple Example
```{r paged.print=FALSE}
syms1 <- c("GOOGL", "AMZN", "AAPL", "MSFT", "CSCO", "INTC", "CVX", "XOM", "SLB",
  "COP", "JPM", "WFC", "USB", "AXP", "WMT", "TGT", "HD", "COST")
df <- sp500_px %>%
  filter(Date >= "2011-01-01") %>%
  select(all_of(syms1)) %>%
  t() # take transpose: to cluster companies, we need the stocks along the rows

d <- dist(df)
hcl <- hclust(d)
```
#### The Dendrogram
```{r paged.print=FALSE}
plot(hcl)
```
```{r paged.print=FALSE}
plot(hcl, ylab = "distance", xlab = "", sub = "", main = "")
```
```{r paged.print=FALSE}
cutree(hcl, k = 4)
```
### Model-Based Clustering
#### Mixtures of Normals
```{r paged.print=FALSE}
df <- sp500_px %>%
  filter(Date >= "2011-01-01") %>%
  select(c(XOM, CVX))
mcl <- Mclust(df)
summary(mcl)
```
```{r paged.print=FALSE}
cluster <- factor(predict(mcl)$classification)
ggplot(df, aes(x = XOM, y = CVX, color = cluster, shape = cluster)) +
  geom_point(alpha = 0.8) +
  scale_shape_manual(
    values = c(1, 3),
    guide = guide_legend(override.aes = aes(size = 2)))
```
```{r paged.print=FALSE}
summary(mcl, parameters = TRUE)$mean
```
```{r paged.print=FALSE}
summary(mcl, parameters = TRUE)$variance
```
#### Selecting the Number of Clusters
```{r paged.print=FALSE}
plot(mcl, what = "BIC", ask = FALSE)
```
```{r paged.print=FALSE}
plot(mcl, what = "BIC", ask = FALSE)
```
### Scaling and Categorical Variables
#### Scaling the Variables
```{r paged.print=FALSE}
loan_data <- read_csv(file.path(DATA_DIR, "loan_data.csv.gz"), show_col_types = FALSE) %>%
  mutate(
    across(where(is.character), as.factor),
    outcome = factor(outcome, levels = c("paid off", "default"))
  )
defaults <- loan_data %>%
  filter(outcome == "default") %>%
  select(c(loan_amnt, annual_inc, revol_bal, open_acc, dti, revol_util))
km <- kmeans(defaults, centers = 4, nstart = 10)
centers <- data.frame(size = km$size, km$centers)
round(centers, digits = 2)
```
```{r paged.print=FALSE}
df0 <- scale(df)
km0 <- kmeans(df0, centers = 4, nstart = 10)
centers0 <- scale(km0$centers, center = FALSE,
  scale = 1 / attr(df0, "scaled:scale"))
centers0 <- scale(centers0, center = -attr(df0, "scaled:center"),
  scale = FALSE)
centers0 <- data.frame(size = km0$size, centers0)
round(centers0, digits = 2)
```
#### Dominant Variables
```{r paged.print=FALSE}
syms <- c("GOOGL", "AMZN", "AAPL", "MSFT", "CSCO", "INTC", "CVX", "XOM", "SLB",
  "COP", "JPM", "WFC", "USB", "AXP", "WMT", "TGT", "HD", "COST")
top_sp1 <- sp500_px %>%
  filter(Date >= "2011-01-01") %>%
  select(all_of(syms))
sp_pca1 <- princomp(top_sp1)
screeplot(sp_pca1)
```
```{r paged.print=FALSE}
components <- 1:length(sp_pca1$sdev)
g <- tibble(
  Components = components,
  Variances = sp_pca1$sdev ** 2,
) |>
filter(
  Components < 11,
) |>
  ggplot(aes(x=Components, y=Variances)) +
    geom_bar(stat="identity") +
    coord_cartesian(xlim = c(0.5, 10.49)) +
    scale_x_continuous(breaks = components) +
    theme_bw()
g
```
```{r paged.print=FALSE}
round(sp_pca1$loadings[, 1:2], 3)
```
#### Categorical Data and Gower's Distance
```{r paged.print=FALSE}
x <- loan_data[1:5, c("dti", "payment_inc_ratio", "home_", "purpose_")]
x
```
```{r paged.print=FALSE}
daisy(x, metric = "gower")
```
```{r paged.print=FALSE}
set.seed(1234)
defaults <- loan_data %>%
  filter(outcome == "default")
df <- defaults[sample(nrow(defaults), 250),
  c("dti", "payment_inc_ratio", "home_", "purpose_")]
d <- daisy(df, metric = "gower")
hcl <- hclust(d)
dnd <- as.dendrogram(hcl)
plot(dnd, leaflab = "none", ylab = "distance")
```
```{r paged.print=FALSE}
plot(dnd, leaflab = "none", ylab = "distance")
```
```{r paged.print=FALSE}
dnd_cut <- cut(dnd, h = 0.5)
df[labels(dnd_cut$lower[[1]]), ]
```
#### Problems with Clustering Mixed Data
```{r paged.print=FALSE}
df <- model.matrix(~ -1 + dti + payment_inc_ratio + home_ + pub_rec_zero,
  data = defaults)
df0 <- scale(df)
km0 <- kmeans(df0, centers = 4, nstart = 10)
centers0 <- scale(km0$centers, center = FALSE,
  scale = 1 / attr(df0, "scaled:scale"))
round(scale(centers0, center = - attr(df0, "scaled:center"), scale = FALSE), 2)
```
# Supplementary Material
## Figure 7-4. Graphical representation of a correspondence analysis of house task data
```{r paged.print=FALSE}
set.seed(1234)
contrib <- ca_analysis$sv ** 2
contrib <- contrib / sum(contrib)
colcoord <- as.data.frame(ca_analysis$colcoord)
rowcoord <- as.data.frame(ca_analysis$rowcoord)
coords <- bind_rows(
  rowcoord %>% mutate(type = "rowcoord"),
  colcoord %>% mutate(type = "columns")
)
row.names(coords) <- gsub("_", " ", row.names(coords))

graph <- ggplot(coords, aes(x = Dim1, y = Dim2, color = type,
    label = rownames(coords), shape = type)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "#444444") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "#444444") +
  geom_point() +
  geom_text_repel() +
  xlab(sprintf("Dimension 1 (%.1f%%)", 100 * contrib[1])) +
  ylab(sprintf("Dimension 2 (%.1f%%)", 100 * contrib[2])) +
  scale_color_manual(values = c("blue", "red")) +
  theme_bw() +
  theme(legend.position = "none")
graph
```
## Figure 7-5. The clusters of K-means applied to daily stock returns for ExxonMobil and Chevron
```{r paged.print=FALSE}
set.seed(1010103)
df <- sp500_px %>%
  filter(Date >= "2011-01-01") %>%
  select(XOM, CVX)
km <- kmeans(df, centers = 4, nstart = 1)
df <- df %>% mutate(cluster = factor(km$cluster))
centers <- as.tibble(km$centers) %>%
  mutate(cluster = factor(1:4))

ggplot(df, aes(x = XOM, y = CVX, color = cluster, shape = cluster)) +
  geom_point() +
  scale_shape_manual(
    values = c(1, 3, 2, 4),
    guide = guide_legend(override.aes = aes(size = 1))
  ) +
  geom_point(data = centers, aes(x = XOM, y = CVX), size = 2, stroke = 2, color = "black") +
  coord_cartesian(xlim = c(-2, 2), ylim = c(-2.5, 2.5)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))
```
## Figure 7-9. A comparison of measures of dissimilarity applied to stock data
```{r paged.print=FALSE}
cluster_fun <- function(df, method) {
  d <- dist(df)
  hcl <- hclust(d, method = method)
  tree <- cutree(hcl, k = 4)
  return(df %>% mutate(cluster = factor(tree), method = method))
}

df0 <- sp500_px %>%
  filter(Date >= "2011-01-01") %>%
  select(XOM, CVX)
df <- bind_rows(
  cluster_fun(df0, method = "single"),
  cluster_fun(df0, method = "average"),
  cluster_fun(df0, method = "complete"),
  cluster_fun(df0, method = "ward.D")
)
df$method <- ordered(df$method, c("single", "average", "complete", "ward.D"))

graph <- ggplot(data = df, aes(x = XOM, y = CVX, color = cluster, shape = cluster)) +
  geom_point(alpha = 0.6) +
  scale_shape_manual(
    values = c(1, 3, 4, 2),
    guide = guide_legend(override.aes = aes(size = 2))
  ) +
  facet_wrap(~ method)
graph
```
## Figure 7-10. Probability contours for a two-dimensional normal distribution
```{r paged.print=FALSE}
mu <- c(0.5, -0.5)
sigma <- matrix(c(1, 1, 1, 2), nrow = 2)
prob <- c(0.5, 0.75, 0.95, 0.99) ## or whatever you want
names(prob) <- prob ## to get id column in result
df <- tibble()
for (p in prob){
  df <- bind_rows(
    df,
    as.tibble(ellipse(x = sigma, centre = mu, level = p)) %>%
      mutate(prob = p)
  )
}
names(df) <- c("X", "Y", "Prob")

## Figure 7-9: Multivariate normal ellipses
dfmu <- tibble(X = mu[1], Y = mu[2])

graph <- ggplot(df, aes(X, Y)) +
  geom_path(aes(linetype = factor(Prob))) +
  geom_point(data = dfmu, aes(X, Y), size = 3)
graph
```
